Meeting the Challenges of Modeling Astrophysical Thermonuclear Explosions: \castro, \maestro, and the \amrex\ Astrophysics Suite

### 2017-10-21 astronum_2017/astronum-amrex.tex
update with Brian's comments

\author{M. Zingale$^1$,
        A.~S. Almgren$^2$,
        M.~G. Barrios Sazo$^1$,
        V.~E. Beckner$^2$,
        J.~B. Bell$^2$,
        B. Friesen$^3,2$,
        A.~M. Jacobs$^4$,
        M.~P. Katz$^5$,
        C.~M. Malone$^6$,
        A.~J. Nonaka$^2$,
        D. Willcox$^1$, and
        W. Zhang$^2$}

@@

Astrophysical explosions come in many flavors: gravitational and
thermonuclear supernovae, unstable burning on the surface of compact
objects, and explosive ignition of burning stages in stellar
evolution.  Accurate modeling of these events requires the coupling of
hydrodynamics, gravity, thermonuclear reactions, and in some cases,
radiation and magnetic fields.  Further, these environments are
characterized by a wide range of lengthscales, from the size of the
star or binary system down to the burning zone width and dissipation
scales.  Temporal scales are equally impressive---stellar evolution
occurs over 10s of millions to billions of years, the simmering phase
leading up to explosions lasts millenia to days or hours, and the
explosion can be over in seconds to hours.  The radiation leakage,
which leads to the observables we see lasts from hours to months.

@@

The largest open theoretical question for SNe Ia is---what is the
nature of the progenitor?  About 20 years ago, the community had
mostly converged up the single-degenerate scenario---a
Chandrasekhar-mass C/O white dwarf that accretes from its companion,
eventually leading to runaway at the center that burns through the
star (see~\cite{hillebrandtniemeyer2000} for the state of the field at
that time).  Since then, the wealth of observations has show that
there is a lot of diversity in SNe Ia, and searches for progenitor
systems have strongly suggested that Chandra-mass white dwarfs cannot
explain most SNe Ia.  Today, merging white dwarfs (the double
degenerate scenario) have perhaps become the most popular model.
Other progenitors, like He burning on the surface of a sub-Chandra
white dwarf have also seen interest in explaining some of the observed
diversity.  See~\cite{araa-maoz} for a review.

There are open questions in all of these scenarios that can be
addressed through simulation.  For the Chandra and sub-Chandra models,
what is the distribution (spatial and temporal) of the hotspots setup
by turbulent convection that give rise to burning fronts?  A
longstanding question with the Chandra model is, can a deflagration
transition to a detonation during the burning front propagation
through the star.  For the sub-Chandra double detonation model, is it
possible to create a detonation in the thin surface He layer?  For
double degenerates, does the burning take place promptly? Can we
avoid an accretion-induced collapse to a neutron star?  For many of
these investigations, we need to address numerical issues such as 
whether it is
possible to accurately model the ignition of a detonation with the
spatial resolution we can attain.  These are some of the questions we
seek to answer.

@@

Our suite of application codes is built on the
\amrex\ block-structured adaptive mesh refinement (AMR) library.  The
basic programming model has \amrex\ managing the grid data-structures
and parallel communications, and it calls the computational kernels on
a patch-by-patch basis.  The core library is written in \cpp\ with 
computational kernels written in Fortran\footnote{Currently
\maestro\ is written in pure-Fortran, but will be
 ported to the updated \cpp\ \amrex\ framework this coming year.}
---this allows us to take
advantage of the strengths of both languages. \amrex\ supports subcycling
in time, which we use in \castro.
\begin{figure}[t]
\centering
\includegraphics[width=0.58\linewidth]{slice_grid}
\begin{minipage}[b]{0.38\linewidth}
\caption{\label{fig:amr_grid} A slice through a
  \castro\ three-dimensional simulation of merging white dwarfs
  showing the 2-level grid structure.  The boxes making up the base
  grid are drawn in white and the boxes making up the finer grid are drawn in black.  Note
  that the finer boxes can overlap multiple coarse boxes.\vspace{2em}}
\end{minipage}
\end{figure}

\amrex\ uses a hybrid MPI + OpenMP approach to parallelism.
Distribution of grid patches to nodes using MPI provides a natural
coarse-grained approach to distributing the computational work, while
subdividing patches into tiles and using threads to parallelize over
tiles using OpenMP provides effective fine-scale parallelization and
amortizes thread overhead over large units of work.  Additionally,
tiling does not incur the large increase in metadata
associated with using smaller patches in a flat MPI
mode~\cite{tiling} since we can use fewer, larger patches.
This strategy is especially important for
many-core architectures like the Intel Phi.  Ongoing development is being
done to support GPU offloading, using managed memory provided by the
latest generations of GPUs.

@@

A key design goal of our application codes is performance portability.
We want the same kernels to run on clusters, manycore machines (e.g.
Intel KNL), and GPU-based machines.  Our development has balanced this
need with architecture-specific optimizations to maximize code reuse.

\begin{figure}[t]
\centering
\includegraphics[width=0.48\linewidth]{titan_edison_maestro_scaling}
\begin{minipage}[b]{0.48\linewidth}
\caption{\label{fig:maestro_scaling} \maestro\ strong scaling on NERSC
  edison and OLCF titan for a 3-d XRB problem.  Two different problem
  sizes are shown.  We see excellent strong scaling to high core counts
  for this problem.\vspace{2em}}
\end{minipage}
\end{figure}

\begin{figure}[t]
\centering
\includegraphics[width=0.48\linewidth]{sedov_scaling}
\includegraphics[width=0.48\linewidth]{wdmerger_scaling}
\caption{\label{fig:castro-scaling} (left) \castro\ strong scaling on
  OLCF titan for a pure hydro (Sedov w/ real EOS) problem.  The
  different colors represent different base resolutions.  We see
  excellent strong and weak scaling for the single level runs.  For
  the $1024^3$ run, we also ran with one level of refinement by a
  factor of 2 (triangles) or a factor of 4 (squares), and see good
  strong scaling.  The variability (shown by the error bars) at high
  core counts shows we are becoming work-starved. (right)
  \castro\ strong scaling on OLCF titan for a hydro + Poisson gravity
  (wdmerger) problem.  This problem uses a multipole solver to
  determine Dirichlet boundary conditions representing an isolated
  mass distribution, and then geometric multigrid to solve for the
  potential in the interior.  Two coarse grid sizes are shown, and
  demonstrate great strong scaling.  We also look at a single level of
  refinement (by a factor of 4) on top of this coarse grid.}
\end{figure}

\begin{figure}[t]
\centering
\includegraphics[width=0.48\linewidth]{summitdev_scaling}
\includegraphics[width=0.48\linewidth]{cori_scaling}
\caption{\label{fig:knl-gpu-scaling} \castro\ scaling on new
  architectures.  The left figure shows the \castro\ GPU proxy app
  (\starlord) on OLCF summitdev.  We see good scaling behavior to 64
  GPUs in this test of a pure hydro problem (Sedov w/ real EOS).  On the
right, we explore performance on NERSC cori (and Intel KNL platform).}
\end{figure}

Figure~\ref{fig:maestro_scaling} shows strong scaling for \maestro\ on
the 3-d XRB problem.  This is a typical
\maestro\ application~\cite{xrb3}, where burning can be a significant
part of the overall evolution.  We ran on both OLCF titan and NERSC
edison.  We see that the code scales well to $\mathcal{O}(10^4)$
processors.  The upturn at the end of the scaling curves on titan
reflect the change from 1 MPI task / 8 OpenMP threads per NUMA node to
1 MPI task / 16 OpenMP threads per compute node (2 NUMA nodes).  The
main limitation to the scaling at the moment is the multigrid solves
used to enforce the projection (in particular the nodal solver).
Also, \maestro\ currently uses a simpler fine-grained approach to parallelism
where planes in the $z$-direction are divided among OpenMP threads.
As we port \maestro\ to the \cpp\ \amrex\ code base we will take
advantage of ongoing development for increased tile size control.

Figure~\ref{fig:castro-scaling} shows \castro\ scaling behavior for a pure
hydro + real EOS problem and a hydro / real EOS, + self-gravity
problem.  All tests were run on OLCF titan.  The AMD processors in
titan can be used in an 8 or 16 cores / node configuration.  For all
these runs, we ran in the 16 cores / node mode.  All runs also used
the PGI 17.7 compilers.  We see excellent scaling behavior for the
pure hydro problem.  For the self-gravity problem, we scale well with
a uniform grid, but with a factor of 4 refinement for the AMR level on
top of this, we observe degraded performance at higher core counts.
This is a challenging problem, as only 3\% of the domain volume is
refined.  For a given core count, there are a number of different
combinations of MPI ranks and OpenMP threads we can use.  In general,
with multilevel problems, we found the best performance with fewer MPI
ranks and more OpenMP threads.

@@

Figure~\ref{fig:knl-gpu-scaling} shows the performance of
\starlord\ on the summitdev platform at OLCF\footnote{summitdev
  consists of 2 IBM Power8 processors and 4 Nvidia Pascal GPUs per
  node.}.  The highest GPU count (128) corresponds to 32 nodes on
summitdev.  We see a nearly linear speedup with the number of GPUs,
indicating good weak scaling on this machine.  Efforts are underway to
complete the port of the GPU developments into \castro.  This figure
also shows our performance on the Intel KNL manycore processors (using
NERSC cori).  This is for the same problem as the self-gravity test on
titan. \MarginPar{need more info: base grid?}  The KNL chips in cori
have 68 cores than can be run with 1, 2, or 4 hardware threads each.
For all these runs, we ran with 4 threads / core (effectively 272
threads / chip).

@@

>>>>

\author{M. Zingale$^1$,
        A.~S. Almgren$^2$,
        M.~G. Barrios Sazo$^1$,
        V.~E. Beckner$^2$,
        J.~B. Bell$^2$,
        B. Friesen$^{3,2}$,
        A.~M. Jacobs$^4$,
        M.~P. Katz$^5$,
        C.~M. Malone$^6$,
        A.~J. Nonaka$^2$,
        D. Willcox$^1$, and
        W. Zhang$^2$}

@@

Astrophysical explosions come in many flavors: gravitational and
thermonuclear supernovae, unstable burning on the surface of compact
objects, and explosive ignition of burning stages in stellar
evolution.  Accurate modeling of these events requires the coupling of
hydrodynamics, gravity, thermonuclear reactions, and in some cases,
radiation and magnetic fields.  Further, these environments are
characterized by a wide range of length scales, from the size of the
star or binary system down to the burning zone width and dissipation
scales.  Temporal scales are equally impressive---stellar evolution
occurs over 10s of millions to billions of years, the simmering phase
leading up to explosions lasts hours or days to millenia, and the
explosion can be over in seconds to hours.  The radiation leakage,
which leads to the observables we see lasts from hours to months.

@@

Among the most significant open questions for SNe Ia is the identity of the
progenitor.  About 20 years ago, the community had
mostly converged upon the single-degenerate scenario---a
Chandrasekhar-mass C/O white dwarf that accretes from its companion,
eventually leading to runaway at the center that burns through the
star (see~\cite{hillebrandtniemeyer2000} for the state of the field at
that time).  Since then, the wealth of observations has indicated that
there is a lot of diversity in SNe Ia, and searches for progenitor
systems have strongly suggested that Chandra-mass white dwarfs cannot
explain most SNe Ia.  Today, merging white dwarfs (the double
degenerate scenario) have perhaps become the most popular model.
Other progenitors, like He burning on the surface of a sub-Chandra
white dwarf, have also seen interest in explaining some of the observed
diversity.  See~\cite{araa-maoz} for a review.

There are open questions in all of these scenarios that can be
addressed through simulation.  For the Chandra and sub-Chandra models,
what is the distribution (spatial and temporal) of the hotspots setup
by turbulent convection that give rise to burning fronts?  A
longstanding question with the Chandra model is whether a deflagration
can transition to a detonation during the burning front propagation
through the star.  For the sub-Chandra double detonation model, is it
not clear whether it is possible to create a detonation in the thin
surface He layer.  For double degenerates, it is still unresolved
whether the burning takes place promptly or after some delay.  However
it proceeds, we need to avoid an accretion-induced collapse to a
neutron star.  For many of these investigations, we need to address
numerical issues such as whether it is possible to accurately model
the ignition of a detonation with the spatial resolution we can
attain.  These are some of the questions we seek to answer.

@@

Our suite of application codes is built on the
\amrex\ block-structured adaptive mesh refinement (AMR) library.  The
basic programming model has \amrex\ managing the grid data-structures
and parallel communications, and it calls the computational kernels on
a patch-by-patch basis.  The core library is written in \cpp\ with 
computational kernels written in Fortran\footnote{Currently
\maestro\ is written in pure-Fortran, but will be
 ported to the updated \cpp\ \amrex\ framework this coming year.}
---this allows us to take
advantage of the strengths of both languages. \amrex\ supports subcycling
in time, which we use in \castro.
\begin{figure}[t]
\centering
\includegraphics[width=0.58\linewidth]{slice_grid}
\begin{minipage}[b]{0.38\linewidth}
\caption{\label{fig:amr_grid} A slice through a
  \castro\ three-dimensional simulation of merging white dwarfs
  showing the 2-level grid structure.  The boxes making up the base
  grid are drawn in white and the boxes making up the finer grid are drawn in black.  Note
  that the finer boxes can overlap multiple coarse boxes, and not all
  boxes are the same size.\vspace{2em}}
\end{minipage}
\end{figure}

\amrex\ uses a hybrid MPI + OpenMP approach to parallelism.
Distribution of grid patches to nodes using MPI provides a natural
coarse-grained approach to distributing the computational work, while
subdividing patches into tiles and using threads to parallelize over
tiles using OpenMP provides effective fine-scale parallelization and
amortizes thread overhead over large units of work.  Additionally,
tiling does not incur the large increase in metadata
associated with using smaller patches in a flat MPI
mode~\cite{tiling} since we can use fewer, larger patches.
This strategy is especially important for
many-core architectures like the Intel Xeon Phi.  Ongoing development is being
done to support GPU offloading, using managed memory provided by the
latest generations of GPUs.  Figure~\ref{fig:amr_grid} shows an
example of a 2-level grid.

@@

A key design goal of our application codes is performance portability.
We want the same kernels to run on clusters, manycore machines (e.g.
Intel Xeon Phi), and GPU-based machines.  Our development has balanced this
need with architecture-specific optimizations to maximize code reuse.

\begin{figure}[t]
\centering
\includegraphics[width=0.48\linewidth]{titan_edison_maestro_scaling}
\begin{minipage}[b]{0.48\linewidth}
\caption{\label{fig:maestro_scaling} \maestro\ strong scaling on NERSC
  Edison and OLCF Titan for a 3-d XRB problem.  Two different problem
  sizes are shown.  We see excellent strong scaling to high core counts
  for this problem.\vspace{2em}}
\end{minipage}
\end{figure}

\begin{figure}[t]
\centering
\includegraphics[width=0.48\linewidth]{sedov_scaling}
\includegraphics[width=0.48\linewidth]{wdmerger_scaling}
\caption{\label{fig:castro-scaling} (left) \castro\ strong scaling on
  OLCF Titan for a pure hydro (Sedov w/ real EOS) problem.  The
  different colors represent different base resolutions.  We see
  excellent strong and weak scaling for the single level runs.  For
  the $1024^3$ run, we also ran with one level of refinement by a
  factor of 2 (triangles) or a factor of 4 (squares), and see good
  strong scaling.  The variability (shown by the error bars) at high
  core counts shows we are becoming work-starved. (right)
  \castro\ strong scaling on OLCF Titan for a hydro + Poisson gravity
  (wdmerger) problem.  This problem uses a multipole solver to
  determine Dirichlet boundary conditions representing an isolated
  mass distribution, and then geometric multigrid to solve for the
  potential in the interior.  Two coarse grid sizes are shown, and
  demonstrate great strong scaling.  We also look at a single level of
  refinement (by a factor of 4) on top of this coarse grid.}
\end{figure}

\begin{figure}[t]
\centering
\includegraphics[width=0.48\linewidth]{summitdev_scaling}
\includegraphics[width=0.48\linewidth]{cori_scaling}
\caption{\label{fig:knl-gpu-scaling} \castro\ scaling on new
  architectures.  The left figure shows the \castro\ GPU proxy app
  (\starlord) on OLCF summitdev.  We see good scaling behavior to 64
  GPUs in this test of a pure hydro problem (Sedov w/ real EOS).  On the
right, we explore performance on NERSC Cori (and Intel Xeon Phi platform).}
\end{figure}

Figure~\ref{fig:maestro_scaling} shows strong scaling for \maestro\ on
the 3-d XRB problem.  This is a typical
\maestro\ application~\cite{xrb3}, where burning can be a significant
part of the overall evolution.  We ran on both OLCF Titan and NERSC
Edison.  We see that the code scales well to $\mathcal{O}(10^4)$
processors.  The upturn at the end of the scaling curves on Titan
reflect the change from 1 MPI task / 8 OpenMP threads per NUMA node to
1 MPI task / 16 OpenMP threads per compute node (2 NUMA nodes).  The
main limitation to the scaling at the moment is the multigrid solves
used to enforce the projection (in particular the nodal solver).
Also, \maestro\ currently uses a simpler fine-grained approach to parallelism
where planes in the $z$-direction are divided among OpenMP threads.
As we port \maestro\ to the \cpp\ \amrex\ code base we will take
advantage of ongoing development for increased tile size control.

Figure~\ref{fig:castro-scaling} shows \castro\ scaling behavior for a pure
hydro + real EOS problem and a hydro / real EOS, + self-gravity
problem.  All tests were run on OLCF Titan.  The AMD processors in
Titan can be used in an 8 or 16 cores / node configuration.  For all
these runs, we ran in the 16 cores / node mode.  All runs also used
the PGI 17.7 compilers.  We see excellent scaling behavior for the
pure hydro problem.  For the self-gravity problem, we scale well with
a uniform grid, but with a factor of 4 refinement for the AMR level on
top of this, we observe degraded performance at higher core counts.
This is a challenging problem, as only 3\% of the domain volume is
refined.  For a given core count, there are a number of different
combinations of MPI ranks and OpenMP threads we can use.  In general,
with multilevel problems, we found the best performance with fewer MPI
ranks and more OpenMP threads.

@@

Figure~\ref{fig:knl-gpu-scaling} shows the performance of
\starlord\ on the summitdev platform at OLCF\footnote{summitdev
  consists of 2 IBM Power8 processors and 4 Nvidia Pascal GPUs per
  node.}.  The highest GPU count (128) corresponds to 32 nodes on
summitdev.  We see a nearly linear speedup with the number of GPUs,
indicating good weak scaling on this machine.  Efforts are underway to
complete the port of the GPU developments into \castro.  This figure
also shows our performance on the Intel Xeon Phi manycore processors (using
NERSC Cori).  This is for the same problem as the self-gravity test on
Titan.  The Intel Xeon Phi chips in Cori
have 68 cores than can be run with 1, 2, or 4 hardware threads each.
For all these runs, we ran with 4 threads / core (272
threads / chip).

@@

\MarginPar{we should update this to a more modern processor---Don can you run
this on summitdev?}
